---
title: "Invasive Herpetofauna"
author: "Gavin Masterson"
date: "29/11/2019"
output: github_document
always_allow_html: true
---

```{r global options, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Invasive Herpetofauna

How many species of alien invasive herpetofauna are there globally? What are they? Where are they from? Where are they invasive? Keeping up to date with global trends and situations in herpetofaunal conservation is an ongoing challenge. Some of the most infamous invasive species are reptiles - think of the Burmese Python (*Python bivittatus*) in Florida, or the Brown Tree Snake (*Boiga irregularis*) in Guam - and amphibians - think of the Cane Toad (*Rhinella marina*) in Australia or the Common Platanna (*Xenopus laevis*) in Europe. As a herpetologist, I wanted to develop a reproducible webscraping method that will allow me to keep up to date on alien/invasive herpetofaunal species around the world.

[The Global Invasive Species Database](http://www.iucngisd.org/gisd/) (GISD) is a product of the IUCN's Species Survival Commission. Initial development of the GISD took place between 1998 and 2000, and the database received a functionality/cosmetic upgrade in 2004. The GISD database is an ongoing project and is curated by the work of specialists who volunteer their valuable time and expertise. More information can be found [here](http://www.iucngisd.org/gisd/about.php) on the GISD website.

For the work presented below, I was inspired by [this tutorial](https://naturaldatasolutions.com/2018/11/28/scraping-and-visualizing-the-gisd-with-r/) by Jim Sheehan on November 28, 2018. Jim's work facilitated this - my first webscraping project - and I relied on his post to get me started with the layout of the project workflow. I am extremely grateful to Jim for his post.

## Setting up

As always the first thing to do is to load the list of packages required for the project. This list usually gets built over time as you go through the project and then ends up complete by the end of the work.

```{r setup, message = FALSE}
library(robotstxt)    # check website scraping rules
library(tidyverse)    # simultaneously load the stringr, dplyr, ggplot2, purrr and readr packages
library(rvest)        # web scraping
library(knitr)        # HTML table creation
library(kableExtra)   # additional HTML table formatting
library(plotly)       # interactive plots
library(gganimate)    # animate ggplots
library(rworldmap)    # world chloropleth mapping
```

## Webscraping Invasive Herpetofauna from the GISD

The code below imports the saved .csv file from my Advanced Search on the GISD.I limited my search to Reptiles and Amphibians and saved the search return as a .csv file named "amrep_gisd.csv". After importing the data I noticed that there were two issues. Several entries in the "Order" column are currently missing from the database. The second issue is that the importing process adds an additional column of NA values, which needs to be removed. The last two lines of code in the chunk below generate a table of a random sample of 10 of the species. The table is just to give us a quick glance of what the data look like, so the presentation order of the species is not important in this case.

```{r import, warning = FALSE, message = FALSE}

sp_list <- read_delim("amrep_gisd.csv", trim_ws = TRUE, delim = ";") %>% .[,-8] # import .csv and remove X8 column of "NA" values

kable(sp_list[sample(nrow(sp_list),10),]) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Herpetofaunal taxonomy is quirky because of the historical use of subspecies. This means that we need to allow for the possibility of species whose names contain a *generic name*, *specific epithet* and a *subspecific epithet* in the "Species" column of our dataset. For example, the Red-eared Slider is a widespread invasive that is known by the Latin name of *Trachemys scripta elegans*, where the subspecific epithet is "elegans". The problem here is that whenever a species has no *subspecific epithet*, the code below will add "+NA" to the end of the URL rendering it useless for accessing the page of that species.

```{r URLs, warning = FALSE}

sp_list <- sp_list %>% 
              separate(Species, c("Genus", "Specific_Epithet", "Subspecific_Epithet"),
                       sep = " ", 
                       remove = FALSE) %>% 
              unite(c(Genus, Specific_Epithet, Subspecific_Epithet), 
                    col = "genspp", 
                    sep = "+",
                    remove = FALSE) %>% 
              mutate(url = paste0("http://www.iucngisd.org/gisd/speciesname/", genspp)) %>% 
              select(-genspp)

sp_list[sample(nrow(sp_list),5),]$url  # display a random sample of 5 URLs
```

To fix the URLs we need to snip the ends of the URLS that end in exactly "+NA". The code below has the additional step for removing "+NA" from the URLs of species that do not have a subspecific name. (Note: if you are using R in a language other than English, you should refer to documentation for the 'locale'argument of the `coll()` function in `stringr`.)

```{r fixed URLs}

sp_list$url <- str_replace(sp_list$url, pattern = coll(pattern = "+NA"), replacement = "")

sp_list[c(1, 15, 17, 19, 40),]$url  # display five selected URLs
```

The URLs are now correctly specified and we can now scrape each of the species reports. As mentioned above, I am interested in keeping up-to-date with the total number of invasive herpetofauna, where they are invasive and where they are from. The total number can already be deduced from the length of the `sp_list` object. Now we need to scrape the lists of countries where each species is native and where they have established. This involves looping across each URL in `sp_list` and selecting the correct HTML nodes to populate the two, pre-specified lists that are created to receive the scraped information. I have modified the object names, have not used `tryCatch` and I have directly assigned each scrape into the relevant list.

```{r scrape}
urls <- pull(sp_list, url)
sp_names <- pull(sp_list, Species)
a_range <- vector(mode = "list", length = length(urls))
n_range <- vector(mode = "list", length = length(urls))

names(a_range) <- sp_names 
names(n_range) <- sp_names

# Scrape each URL with a 5-second delay in between each iteration

for(i in 1:length(urls)){
    a_range[[i]] <-  read_html(urls[i]) %>% 
                html_nodes(xpath = '//*[@id="l-1st-step"]') %>% 
                html_nodes("li") %>% 
                html_text()
    
    n_range[[i]] <-  read_html(urls[i]) %>% 
                html_nodes(xpath = '//*[@id="nr-col"]') %>% 
                html_nodes("li") %>% 
                html_text()
    
    Sys.sleep(5)
}
rm(i)
```

The next step is to build the tibbles containing either the native range or alien range data in tidy format. Due to the differences in the number of native range and the alien range countries for each species, merging the two tibbles while the countries are specified individually creates readability issues that I would rather sidestep here.

### Countries where the species is indigenous

```{r native range}
n_range_df <- data.frame(Species = rep(names(n_range), 
                                       sapply(n_range, length)),
                         Native_range = unlist(n_range), 
                         row.names = NULL, 
                         stringsAsFactors = FALSE)

n_range_tbl <- inner_join(sp_list, n_range_df, 
                          by = "Species") %>%
               select(Species, Native_range, everything()) %>% 
               as_tibble()
```

### Countries where the species has established an alien or invasive population

```{r alien range}
a_range_df <- data.frame(Species = rep(names(a_range), 
                                       sapply(a_range, length)),
                         Alien_range = unlist(a_range), 
                         row.names = NULL, 
                         stringsAsFactors = FALSE)

a_range_df$Alien_range <- str_replace(a_range_df$Alien_range, 
                                      pattern = "\\[[:digit:]+\\]\\s?", 
                                      "")

a_range_tbl <-  inner_join(sp_list,
                           a_range_df,
                           by = "Species") %>% 
                select(Species, Alien_range, everything()) %>% 
                as_tibble()
```

# Analysis
## Welcome to the world of the #HerpMafia

Closely-related species are known to share biological, behavioural and ecological characteristics. This is important in invasion science because when one species establishes an invasive population it is a good rule-of-thumb to treat closely-related species with extreme caution. So let's start with a look at:
1. which herpetofaunal families have already demonstrated invasive potential and how often? 
2. how many species in each family have established populations outside their natural range? 
3. how many total unique invasions each family is responsible for?
4. which countries are the currently the most impacted by the #HerpMafia?

### 1. Who comprises the #HerpMafia?

The #HerpMafia currently consists of 43 species, from 18 families. There are six families of anuran amphibians (i.e. frogs), but no salamanders (yet?). The most infamous invasive amphibian species to date are the African Clawed Frog *Xenopus laevis* and the Cane Toad *Rhinella marina*. There are 12 reptile families, representing crocodiles, lizards and snakes. The most infamous invasive species of reptile are the Red-eared Slider *Trachemys scripta elegans* (Worldwide), the Burmese Python *Python bivittatus* (Everglades, Florida, USA) and the Brown Tree Snake *Boiga irregularis* (Guam). Thankfully, there are no records of any of the cobra (e.g. Elapidae) or viper (e.g. Viperidae, Crotalidae) families in the GISD. Snakebite is [**already enough of a burden**](https://www.who.int/snakebites/resources/s40409-017-0127-6/en/) in countries with indigenous populations of these snakes. Having to identify bites from, and source venom for an alien invasive venomous snake would be a medical and logistical nightmare with potentially serious outcomes for envenomated humans. 

This year, Ireland recorded its [first ever hospitalisation for a venomous snakebite](https://www.iol.co.za/news/world/ireland-reports-the-first-venomous-snake-bite-in-its-history-right-before-st-patricks-day-44239335). The bite was the outcome of importing a Puff Adder (*Bitis arietans*) as a pet. The necessary antivenom was not available in Ireland and had to be shipped in from Liverpool, UK to assist in the treatment. The irony is that the bite happened just before St Patrick's Day - the day which celebrates the saint who 'drove all the snakes out of Ireland'. If the import of venomous snakes increases, Ireland may require St Patrick's skills again in the near future.

### 2. How many species represent for each #HerpMafia family?

Eleven of the #HerpMafia families are represented by exactly one species. This makes me wonder if they can be called 'families' in the #HerpMafia context... but let us not overlook them because as we'll see next - all it takes is one!

Of the seven families for which more than one species has established an alien range population, five are reptiles and two are amphibians. The family with the most representatives is the Polychrotidae, with 14 *Anolis* species and two *Norops* species. Technically speaking these species have been reassigned to the Dactyloidae family by a recent taxonomic revision - possibly so they can try to dodge the invasion police! See [my {taxadb} blog bost](https://gavinmasterson.netlify.app/post/taxadb/) for ways to check for and resolve taxonomic revisions in data like these here.

```{r represent, warning = FALSE}

sp_fam <- a_range_tbl %>%
          mutate(Family = as_factor(Family)) %>% 
          group_by(Family) %>% 
          tally(n_distinct(Species), name = "Total_species")

#NEED TO LOOK AT WHETHER THIS IS WORTH INCLUDING - IT MEANS THAT THINGS CAN BREAK IF YOU RERUN THIS SECTION OF CODE AND NOT THE WHOLE CHUNK

(sp_fam <- inner_join(x = filter(unique(sp_list[,c(7,9)])), y = sp_fam, by = "Family") %>% 
          mutate(Family = as_factor(str_replace(Family, 
                                      pattern = coll("Polychrotidae"), 
                                      replacement = "Dactyloidae"))) %>% 
          arrange(desc(Total_species)))

```

### 3. How many unique invasions can be linked to each #HerpMafia family?

The code below counts the number of rows for each family and then generates an interactive ggplot of the result. (Try mousing over the bars!) I call the measure "total unique invasions" because each unique species + country pair counts towards the total, even if the same country is invaded by multiple species from the same #HerpMafia family.

```{r unique invasion, warning = FALSE}

inv_family <- a_range_tbl %>%
              mutate (Family = str_replace(Family, 
                                           pattern = coll("Polychrotidae"), 
                                           replacement = "Dactyloidae")) %>% 
              group_by (Family) %>% 
              count(name = "Total_invasions")

inv_family <- inner_join(x = sp_fam, y = inv_family, by = "Family") %>% 
              mutate(Class = as_factor(Class),
                     Family = as_factor(Family)) %>% 
              arrange(desc(Total_invasions))
              
invfam_plot <-  ggplot(inv_family) + 
                geom_col(mapping = aes(x = reorder(Family, -Total_invasions), 
                                       y = Total_invasions,
                                       fill = Class,
                                       text = paste('Family:', Family,
                                                    '<br> Total Unique Invasions:', Total_invasions))) +
                scale_fill_manual(values =c("darkgoldenrod1","chartreuse3")) +
                ylab("Total number of Unique Invasions") +
                xlab("Family") +
                ggtitle("Total Unique Invasions by each #HerpMafia Family") +
                theme(legend.position = "right", 
                      legend.title = element_blank(),
                      axis.text.x = element_text(angle = 90, hjust = 1, vjust = -0.3))

ggplotly(invfam_plot, tooltip = "text")
```

The graph above is simple enough to interpret but for a clearer understanding of herpetofaunal invasions we need to add the information about each family's species tally. Let's do that now.

```{r two plus three}

(XYplot <- ggplot(inv_family) +
          geom_point(mapping = aes(x = Total_species,
                                   y = Total_invasions,
                                   colour = Class)) +
          labs(title = "Scatterplot of Species vs Unique Invasions",
               x = "Invasive Species per Family",
               y = "Total Unique Invasions"))
```

Despite being outnumbered and outrepresented, the amphibian families punch above their weight in terms of total unique invasions.

## Number of alien range countries for each species


# Appendix

## Bonus Code Snippets

I made an animated version of the Total Unique Invasions bar chart to post to my Twitter account on 25 Feb 2020. This is the code for generating the animated version with the Family names removed from the X-axis. The biggest difference is that you need to add a variable that specifies the states of the chart during animation. In this case I added a variable called "Frame" with two states - 0 invasions or all invasions. I had to create a dataframe with zero invasions for all families and then join them together for the plot. 

```{r animated plot}

a <- data.frame(Family = c(inv_family$Family), Total_invasions = rep(0,18), frame = rep('a',18))
b <- data.frame(Family = c(inv_family$Family), Total_invasions = c(inv_family$Total_invasions), frame = rep('b',18))
invfa_anim_df <- rbind(a,b); rm(a,b)

invfam_anim <- ggplot(invfa_anim_df, aes(x = reorder(Family,-Total_invasions), 
                                         y = Total_invasions, 
                                         fill = reorder(Family,-Total_invasions))) +
                geom_bar(stat = 'identity') + 
                labs(x ="Family",
                     y = "Total number of Unique Invasions",
                     title = "Total Unique Invasions for each #HerpMafia Family") +
                #scale_x_discrete(labels= c(rep("?",18))) +
                theme(legend.position = "none",
                      axis.text.x = element_text(angle = 90, hjust = 1)) +
                transition_states(
                    frame,
                    transition_length = 2,
                    state_length = 1,
                    wrap = TRUE
                )

print(invfam_anim) 

#anim_save("unique_invasions_by_family_ordered.gif", animation = invfam_anim)
```

```{r poll figure}
poll_fig <- ggplot(inv_family) + 
                geom_col(mapping = aes(x = Family, 
                                       y = Total_invasions,
                                       fill = Family,
                                       text = paste('Family:', Family,
                                                    '<br> Total Unique Invasions:', Total_invasions))) +
                ylab("Total number of Unique Invasions") +
                xlab("Family") +
                ggtitle("Total Unique Invasions by each #HerpMafia Family") +
                theme(legend.position = "none", 
                      legend.title = element_blank(),
                      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))

#ggsave(filename = "poll_fig.png", plot = poll_fig, width = 15, height = 10, dpi = 200, units = "cm", device='png')
```


