---
title: "taxize"
author: "Gavin Masterson"
date: "19/03/2020"
output: html_document
---
```{r setup, message = FALSE}
library(taxize)
library(tidyverse)
```

## {taxize}
- introduction
- API call demo
- global names resolver (using species name data from 2006?)

If you love taxonomic information and it at the tip of your fingertips, `taxize` is the package for you. With `taxize` we can perform queries of data downloaded using the API of various taxonomic databases. To demonstrate the similarity between `taxize` and `taxadb`, I extract the ITIS classification information for all the species listed in the `GISD-query` object used above. To reduce the run time of the code chunk, I used the `classification` function to query just six of the 43 species from the `GISD-query`.

```{r taxize, message = TRUE}
tax_NA <- GISD_query %>% 
            filter(!complete.cases(.)) %>% 
            pull(Species)
print(classification(GISD_query$Species[c(1,16,20,26,31,40)], db = 'itis')[1:6])
```

From the returned list we see that four of the six species have a full classification, and two do not. The data returned for *Elaphe guttata* contains just a species rank entry. This suggests that the name is just a synonym, which we know is the truth from our `taxadb` demo above. We also see that *Norops grahami* is not found in the ITIS database, which indicates a name that may be either very recent or disputed. It turns out, it's the [latter](https://en.wikipedia.org/wiki/Norops). And on the bombshell that taxonomic disagreements can get quite heated, we'll leave the GISD dataset and look at one more useful `taxize` function - `gnr_resolve`.

# Out with the old...

When you have an old species name or a list of old species names (or even a list of names that someone else has given you), it can be difficult to know that they are up to date. This is where `gnr-resolve` comes in. I have a list of reptile species names from a herpetofaunal survey I conducted in 2006.

```{r 2006_reptile, message = FALSE}
survey_2006 <- read_csv("rep_survey.csv")
temp <- gnr_resolve(names = survey_2006[[1]])
head(temp)
```

One of the very useful helper functions to `gnr_resolve` is `gnr_datasources`. Calling the function provides a tibble containing a very detailed breakdown of the datasources used to resolve the name/s supplied to `gnr_reolve`. In the code below I use `gnr_datasources` to see which datasources have been updated within the last 150 days. I used the call to `Sys.time` but for reproducibility I have inserted the actual system time in the code below.

```{r updateTime}
x <- format("2020-03-16 17:46:31 SAST") %>% 
      strptime(format = "%Y-%m-%d %H:%M:%OS")

y <- gnr_datasources()$updated_at %>% 
      str_replace(.,"T"," ") %>% 
      str_replace(.,"Z","") %>% 
      format() %>% 
      strptime(format = "%Y-%m-%d %H:%M:%OS")

tibble(Database_Name = gnr_datasources()$title[which(difftime(x,y) < 150)], 
       last_updated =  difftime(x,y)[(which(difftime(x,y) < 150))]) %>% 
      kable() %>% 
      kable_styling() %>% 
      row_spec(row = 2, bold = TRUE, background = "DodgerBlue", color = "white")
```

Of the 12 data sources in the returned tibble, we see that there is a mixture of open source and curated taxonomic collections. We also see that our trusty ITIS database was updated 127 days ago. 

```{r}
temp %>% 
  filter(data_source_title == "The Reptile Database")
  
```